---
title: "Shelter Distribution Data Cleaning and Visualization with R"
author: "Mohammed Amimul Ehasan Shakib"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,   # Adjust the figure width
  fig.height = 3    # Adjust the figure height
)
```

This project aims to analyze shelter distribution data and visualize the results. The steps below explain how the data is cleaned, transformed, and visualized.

#### Load Necessary Libraries


```{r}

library(dplyr) #for data manipulation
library(ggplot2) #for data visualization
library(readxl) # for reading Excel files
library(tidyr) #for data tidying
library(writexl) #for exporting data to Excel format
```

#### Set Working Directory and Read Data

I have set the working directory where data file is located and read the Excel file, specifically the sheet named "Improved & Composite".

```{r}
setwd('C:/Users/SHAKIBM/OneDrive - UNHCR/Shelter distribution R dashboard')

# Read the Excel file and list sheet names
file_path <- "ShelterDistribution.xlsx"
data <- read_excel(file_path,sheet = "Improved & Composite")
```


#### Clean Column Names

I replace spaces in the column names with underscores to make them easier to work with.

```{r}

#add underscore with in column name values
data <- data %>%
  rename_with(~ gsub(" ", "_", .))
#check structure
str(data)

```




#### Add and Populate New Column

I add a new column called assistantTypeNew to categorize the types of assistance provided based on the Assistance_Type column.

```{r}
# Add New Column with Specific Data Type
data <- data %>%
  mutate(assistantTypeNew=as.character(NA))
#check structure after adding new column
str(data)

```


#### Adding and Populating a New Column Assistance_Type Based on Column Assistance_Type

Create a new column assistantTypeNew in the dataset based on the values of the Assistance_Type column. This step helps categorize the types of assistance provided in a more readable format.

```{r}

#Based on One Column Value Add New Value to New Column

data <- data %>%
  mutate(
    assistantTypeNew = case_when(
      Assistance_Type=="MTS"~"Multi Storied Shelter",
      Assistance_Type=="CBS"~"Composite Bamboo Shelter",
      Assistance_Type=="Single Storied Shelter"~"Single Storied Shelter"
    ))
```

#### Identify Families Without House Numbers

Next identify and count the families that do not have a house number.

```{r}
#check null vlaue in column House_Number
FamilyWithoutHouseNumber <- data %>%
  filter(is.na(House_Number))
print(FamilyWithoutHouseNumber)



# Print the number of null values
num_null_values <- nrow(FamilyWithoutHouseNumber)
print(paste("Family without House Number :", num_null_values))



```
 


#### Add Unique ID Column

Want to add a unique identifier to each row to help with tracking and referencing individual records.

```{r }

# Add unique ID column
data <- data %>%
  mutate(unique_id = row_number())
```
 
#### Clean 'Remarks' Column

Target is to standardize the values in the Remarks column, combining similar remarks into broader categories and replacing missing values with "No remarks".

```{r }


# Replace the values in the 'Remarks' column, replace NA to "No remakrs"
data <- data %>%
  mutate(Remarks = case_when(
    Remarks=="Filling Sand-50cft,Brick-132"~ "Filling Sand-50cft,Brick-132",
    Remarks=="HH condition EVI"~ "HH condition EVI",
    Remarks %in% c("Stilt Shelter","Stilt Shelter(One Big Tarpuline In 4 Shelter Coverage)","Stilt Shelter(Tie Down Only Rope in 4 shelter)")~ "Stilt Shelter",
    Remarks %in% c("2 shelter had provided due to 6 Nos of family member and this was an especial request from CiC office to provide 2 shelter due to relocation of existing shelter.",
                   "2 shelter had provided due to 6 Nos of family member and this was an especial request from CiC office to provide 2 shelter due to relocation of existing shelter.",
                   "2 shelter had provided due to 8 Nos of family member",
                   "2 shelter had provided due to 9 Nos of family member","Double Shelter",
                   "Family same but 2 shelters had provided previously due to extended family",
                   "In coordination with UNHCR Shelter Focal and CIC, two separate shelters were provided under this Family for their larger household size (10 members).",
                   "Family same but they have 02 different rooms previously due to extended family",
                   "Family same but they have 02 rooms previously due to extended family","Family same but they have 02 different rooms previously due to extended family",
                   "Family same due to extended family double Shelter provided") ~ "Family same due to extended family double Shelter provided"
    ,    TRUE ~ "No remarks"
  ))

```
 
  
#### Show unique value of Remarks

Show unique value of Remarks column to check updated value

```{r }
# Get distinct values of the 'Remarks' column
distinct_remarks <- data %>%
  distinct(Remarks)
distinct_remarks

```

#### Update 'Remarks' Based on Conditions

I observe there are some rows vulnerable is Yes but Remakrs column empty,want to update the Remarks column for families identified as vulnerable (Vulnerable column) with the remark "HH condition EVI".

```{r }

#update "HH condition EVI" when EVI=="Yes" & Remarks == "No remarks"
data <- data %>%
  mutate(
    Remarks = case_when(
      Vulnerable == "Yes" & Remarks == "No remarks" ~ "HH condition EVI",
      TRUE ~ Remarks
    )     )


```


#### Convert Missing Values in 'Vulnerable' Column

There are some rows with Vulnerable missing, want to replace missing values in the Vulnerable column with "No" to indicate that these families are not considered vulnerable.

```{r }

# Convert "NA" to "No"
data <- data %>%
  mutate(Vulnerable = ifelse(is.na(Vulnerable), "No", Vulnerable))

# Get distinct values of the 'Remarks' column
Unique_value_of_Vulnerable <- data %>%
  distinct(Vulnerable)
print(Unique_value_of_Vulnerable)

```

#### Stacked Bar Chart


Create a bar chart to show the distribution of different types of assistance across districts.


```{r }

#after update check total data of remarks
totalRemarksWiseData <- data %>%
  group_by(Remarks) %>%
  summarise(n())


#Every Assistance type total records and total unique families supported
groupByassistantTypeNew <- data %>%
  group_by( District_Name,assistantTypeNew) %>%
  summarise(
    TotalRecords= n(),
    UniqueFamiliesSupported =n_distinct(House_Number)
  )


# Create a mapping of Camp_Name to integer values
District_name_mapping <- c(
  "Bandarban" = 1,
  "Brahmanbaria" = 2,
  "Chandpur" = 3,
  "Chattogram" = 4,
  "Coxsbazar" = 5,
  "Cumilla" = 6,
  "Feni" = 7,
  "Habiganj" = 8,
  "Khagrachhari" = 9,
  "Kutupalong RC" = 10,
  "Lakshmipur" = 11,
  "Moulvibazar" = 12,
  "Noakhali" = 13,
  "Rangamati" = 14,
  "Sunamganj" = 15,
  "Sylhet" = 16
)

# Add the integer column to the data
data <- data %>%
  mutate(District_Name_int = District_name_mapping[District_Name])

# Group by Camp_Name, assistantTypeNew, and Camp_Name_int, then summarize
groupByassistantTypeNew <- data %>%
  group_by(District_Name, District_Name_int, assistantTypeNew) %>%
  summarise(
    TotalRecords = n(),
    UniqueFamiliesSupported = n_distinct(House_Number)
  ) %>%
  arrange(District_Name_int, assistantTypeNew) # Sort by the integer column and assistantTypeNew


# Convert Camp_Name to a factor with levels in the sorted order
groupByassistantTypeNew <- groupByassistantTypeNew %>%
  mutate(District_Name = factor(District_Name, levels = names(District_name_mapping)[order(District_name_mapping)]))

```
 
 
 
#### Bar Chart


Create bar chart to show the total number of unique families and records for each type of assistance


```{r }

library(ggplot2)
# Create the stacked bar chart
ggplot(groupByassistantTypeNew, aes(x = District_Name, y = UniqueFamiliesSupported, fill = assistantTypeNew)) +
  geom_bar(stat = "identity") +
  labs(title = "District Name wise Distribution of assistantTypeNew",
       x = "District",
       y = "Total Families Supported",
       fill = "Assistant Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values = c("#0072BC", "#8EBEFF","#DCE9FF"))


```
  
#### Pie Chart

Create a pie chart to show the proportion of families identified as vulnerable.

```{r }
# Calculate the distinct count of EVI for each House_Number
distinct_count <- data %>%
  group_by(Vulnerable) %>%
  summarise(distinct_Vulnerable_count = n_distinct(House_Number))

# Create a pie chart
ggplot(distinct_count, aes(x = "", y = distinct_Vulnerable_count, fill = as.factor(Vulnerable))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar(theta = "y") +
  labs(title = "Families Vulnerable count", y = "Count", x = "") +
  theme_void() +
  theme(legend.title = element_blank())+
  geom_text(aes(label = distinct_Vulnerable_count), position = position_stack(vjust = 0.5))+
  scale_fill_manual(values = c( "#0072BC",
                                "#8EBEFF"))


```
  
  
#### Stacked bar chart

This chart visually represents the distribution of different types of shelter assistance and highlights which type has the highest demand, based on the number of unique families and total records.

```{r }

# Calculate the total distribution of unique families and total records
Total_distribution_Unique_families <- data %>%
  group_by(assistantTypeNew) %>%
  summarise( `Distinct Family Count` = n_distinct(House_Number),
             `Total Records`  = n())

# Reshape the data to a long format for plotting
long_data <- Total_distribution_Unique_families %>%
  pivot_longer(cols = c(`Distinct Family Count`, `Total Records`),
               names_to = "Type",
               values_to = "Count")

 
 # Create a bar chart
ggplot(long_data, aes(x = assistantTypeNew, y = Count, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Distribution of Unique Families and Total Records by Assistant Type",
       x = "Assistant Type",
       y = "Count") +
  theme_minimal()+
  # Use custom colors
  scale_fill_manual(values = c("#0072BC", "#8EBEFF"))

```
    
  
  
#### Save the Modified Data

We save the cleaned and modified data to a new Excel file.

```{r }
 # Save the modified dataframe to a new Excel file
output_file_path <- "C:/Users/SHAKIBM/OneDrive - UNHCR/Shelter distribution R dashboard/ShelterDistributionOutput.xlsx"
write_xlsx(data, output_file_path)

# Print success message
print(paste("File saved to", output_file_path))


```
  
   
